---
title: "XR256_XR249_ISG_integration"
author: "Alexander Lercher"
date: "`r Sys.Date()`"
output: github_document
---

## LOAD PACKAGES
```{r load packages}
library(readr)
library(tidyverse)
library(dplyr)
library(pheatmap)
library(ggpubr)
library(grid)
```

## DATA IMPORT AND CLEANUP
```{r data import and cleanup}
list_of_files <- list.files(path = "../XR256_DESeq2/output",
                            recursive = FALSE,
                            pattern = "DEG.tsv",
                            full.names = TRUE)

# filter for files that contain IFNb exactly once
filtered_files <- list_of_files[str_count(list_of_files, "PIC_vs_PBS") == 1]

data <- readr::read_tsv(filtered_files, id = NULL)

data_fpkm <- read.delim("../XR256_DESeq2/output/XR256_FPKM.tsv")
data_fpkm <- data_fpkm %>%
  dplyr::select(matches("^(gene|PBS|PIC)"))
```

## EXTRACT SIGNIFICANT GENES AND USE FPKM CUTOFF
```{r extract significant genes and use FPKM cutoff}
adj_pval_cutoff = 0.05
log2FC_cutoff = 1
FPKM_cutoff = 1

# sort by gene and by log2FC high to low and remove duplicates
data_sig <- data %>%
  arrange(gene_id, desc(log2FoldChange)) %>%
  distinct(gene_id, .keep_all = T) %>%
  dplyr::select(gene_id,log2FoldChange,padj)

# filter for genes meeting pval and log2FC cutoffs
data_sig <- data_sig %>%
  filter(padj < adj_pval_cutoff, log2FoldChange > log2FC_cutoff)

# merge with FPKM data, add mean fpkm and filter for fpkm cutoff
data_sig <- data_fpkm %>%
  dplyr::rename(gene_id = gene) %>%
  left_join(data_sig) %>%
  na.omit() %>%
  mutate(
    mean_PBS  = rowMeans(dplyr::select(., starts_with("PBS"))),
    mean_PIC = rowMeans(dplyr::select(., starts_with("PIC")))
    ) %>%
  mutate(
    mean_all = rowMeans(dplyr::select(., starts_with("mean")))
  )

data_sig <- data_sig %>%
  filter(mean_all > FPKM_cutoff) %>%
  dplyr::select(matches("^(gene|PBS|PIC)"))
```

## GET LIST OF ISGs FROM XR249
```{r get list of ISGs from XR252}
# get list of ISG names and save to file
# get list of ISG names and save to file
data_ISG <- read.delim("input/XR249_ISG_expression_brain.tsv")
data_ISG <- data_ISG %>%
  dplyr::select(matches("^(gene|PBS|IFN)")) %>%
  dplyr::rename_with(~ str_replace(.x, "^PBS", "PBS_CTL"))
```

## CALCULATE GENE SET EXPRESSION ACROSS CONDITIONS
```{r calculate gene set expression across conditions}
# merge IFN induced genes with PIC induced genes
ISG_fpkm <- data_ISG %>%
  left_join(data_sig) %>%
  dplyr::rename(gene = gene_id) %>%
  na.omit()

# make genes rownames delete gene column
ISG_fpkm <- as.data.frame(ISG_fpkm)
rownames(ISG_fpkm) <- ISG_fpkm$gene
ISG_fpkm$gene = NULL

# calculate average ISG FPKM value per sample
ISG_set_expression <- colSums(ISG_fpkm)/nrow(ISG_fpkm)

# write ISG expression values to file
ISG_set_expression_df <- data.frame(
  condition = names(ISG_set_expression),
  ISG_expression = as.numeric(ISG_set_expression),
  row.names = NULL,
  stringsAsFactors = FALSE) %>%
  write_tsv("output/XR256_ISG_set_expression.tsv")
```

## PLOT SUMMARY DATA as BARPLOTS
```{r plot summary data as barplots}
# define metadata/groups
metadata <- data.frame(
  sample_id = colnames(ISG_fpkm),
  condition = c(rep("PBS_CTL", 3), rep("IFNa", 3), 
                rep("PBS", 4), rep("PIC", 4))
)

# combine metadata with data to plot
plot_data <- metadata %>%
  mutate(score = ISG_set_expression[sample_id])

plot_data$condition <- factor(plot_data$condition, levels = c("PBS_CTL", "IFNa", "PBS", "PIC"))

# prepare y axis breaks and labels
max_exp <- ceiling(log2(max(plot_data$score, na.rm = TRUE)))
breaks <- 2^(0:max_exp)
labels <- parse(text = paste0("2^", 0:max_exp))

# prepare labels and position
label_data <- plot_data %>%
  group_by(condition) %>%
  summarise(mean_val = mean(score, na.rm = TRUE),
            y_pos = max(score, na.rm = TRUE) * 1.1)

# plot data
p <- ggplot(plot_data, aes(x = condition, y = score, fill = condition)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, size = 2, alpha = 0.7) +
    geom_text(data = label_data,
              aes(x = condition, y = y_pos, label = round(mean_val, 1)),
              inherit.aes = FALSE,
              size = 3.5,
              color = "black") +
    scale_y_continuous(
      trans = "log2",
      breaks = breaks,
      labels = labels) +
    labs(
      title = "Expression of ISGs",
      y = "Mean FPKM of ISG set",
      x = "") +
    theme_bw() +
    theme(aspect.ratio = 1)
print(p)
```

## CURATE DATA JOIN DATASETS INDUCED BY POLYIC AND IFNA KEEP ALL THAT ARE UP BY POLYIC
```{r curate data join datasets induced by polyIC and IFNa keep all that are up by polyIC}
# merge PIC induced genes with IFNa induced genes but keep all from PIC
data_all_IFNa <- read.delim("../../XR249_TL396_iv_IFN-a_brain_RNAseq/XR249_RNAseq_analyses_AL/XR249_DESeq2/output/XR249_FPKM.tsv")
data_all_IFNa <- data_all_IFNa %>%
  dplyr::rename_with(~ str_replace(.x, "^PBS", "PBS_CTL")) %>%
  rename(gene_id = gene)
  
data_fpkm <- data_sig %>%
  left_join(data_all_IFNa)

# make genes rownames delete gene column
data_fpkm <- as.data.frame(data_fpkm)
rownames(data_fpkm) <- data_fpkm$gene_id
data_fpkm$gene_id = NULL
```

## PLOT HEATMAP OF INDIVIDUAL VALUES GENES UP BY POLYIC
```{r plot heatmap of individual values genes up by polyIC}
# define formula to calculate row Z score
cal_z_score <- function(x){
  (x - mean(x)) / sd(x)
}

# calculate row Z score for FPKMs for each row (gene) of interest
data_fpkm_norm <- t(apply(data_fpkm, 1, cal_z_score))

# annotate sample columns
my_sample_col <- data.frame(sample = c(rep("PBS",4),
                                       rep("PIC",4),
                                       rep("PBS_CTL",3),
                                       rep("IFNa",3)),
                            condition = c(rep("ctl",4),
                                          rep("polyIC",4),
                                          rep("ctl",3),
                                          rep("IFNa",3)))

row.names(my_sample_col) <- colnames(data_fpkm)

# calculate dendrogram
hc <- hclust(dist(data_fpkm_norm), method = "complete")

# add cluster information to heatmap
numberofclusters = 2
my_gene_col <-cutree(hc, k = numberofclusters)
my_gene_col <- data.frame(cluster = ifelse(test = my_gene_col == 1, yes = "1", no = "2"))

# add ISG information to heatmap
ISG_list <- data_ISG$gene_id
my_gene_col$ISG <- ifelse(rownames(my_gene_col) %in% ISG_list, 1, 0)

p <- pheatmap::pheatmap(data_fpkm_norm, annotation_col = my_sample_col, annotation_row = my_gene_col, 
             show_rownames = F, main = "genes induced by poly(I:C)",
             #color = colorRampPalette(c("midnightblue", "ivory", "firebrick3"))(50),
             cutree_rows = numberofclusters)
grid::grid.newpage()
grid::grid.draw(p$gtable)

# add cluster and ISG information to DEG list
my_gene_col$gene_id <- rownames(my_gene_col)
data_fpkm$gene_id <- rownames(data_fpkm)
data_save <- data_fpkm %>%
  left_join(my_gene_col) %>%
  na.omit() %>%
  dplyr::relocate(gene_id) %>%
  arrange(cluster)

write_tsv(data_save, "output/XR259_ISG_heatmap_individual.tsv")
```

## PLOT HEATMAP OF MEAN VALUES GENES UP BY POLYIC
```{r plot heatmap of mean values genes up by polyIC}
# add means for all conditions
# add means for all conditions
data_fpkm_mean <- data_fpkm %>%
  mutate(
    mean_PBS  = rowMeans(select(., starts_with("PBS"))),
    mean_PIC = rowMeans(select(., starts_with("PIC"))),
    mean_PBS_CTL  = rowMeans(select(., starts_with("PBS_CTL"))),
    mean_IFNa = rowMeans(select(., starts_with("IFNa")))
  )

# keep mean FPKM values only
data_fpkm_mean <- data_fpkm_mean %>%
  dplyr::select(contains(c("mean")))    

# calculate row Z score for FPKMs for each row (gene) of interest
data_fpkm_mean_norm <- t(apply(data_fpkm_mean, 1, cal_z_score)) 

# annotate sample columns
my_sample_col <- data.frame(sample = c(rep("PBS",1),
                                       rep("PIC",1),
                                       rep("PBS_CTL",1),
                                       rep("IFNa",1)),
                            condition = c(rep("ctl",1),
                                          rep("polyIC",1),
                                          rep("ctl",1),
                                          rep("IFNa",1)))

row.names(my_sample_col) <- colnames(data_fpkm_mean)

# calculate dendrogram
hc <- hclust(dist(data_fpkm_mean_norm), method = "complete")

# add cluster information to heatmap
numberofclusters = 2
my_gene_col <-cutree(hc, k = numberofclusters)
my_gene_col <- data.frame(cluster = ifelse(test = my_gene_col == 1, yes = "1", no = "2"))

# add ISG information to heatmap
my_gene_col$ISG <- ifelse(rownames(my_gene_col) %in% ISG_list, 1, 0)

# plot heatmap
p <- pheatmap::pheatmap(data_fpkm_mean_norm, annotation_col = my_sample_col, annotation_row = my_gene_col, 
             show_rownames = F, main = "genes induced by poly(I:C)",
             #color = colorRampPalette(c("midnightblue", "ivory", "firebrick3"))(50),
             cutree_rows = numberofclusters
             )
grid::grid.newpage()
grid::grid.draw(p$gtable)

# add cluster and ISG information to DEG list
my_gene_col$gene_id <- rownames(my_gene_col) 
data_fpkm$gene_id <- rownames(data_fpkm)
data_save <- data_fpkm %>%
  left_join(my_gene_col) %>%
  na.omit() %>%
  dplyr::relocate(gene_id) %>%
  arrange(cluster)

write_tsv(data_save, "output/XR256_ISG_heatmap_means.tsv")
```

## APPENDIX
```{r appendix}
# which R packages and versions?
if ("devtools" %in% installed.packages()) devtools::session_info()
```

